Data Specialization for a Staged Lambda Calculus

### 2015-02-26 paper/splitting.tex
More preample to 4.0.

In this section, we present a \emph{splitting algorithm} which
statically converts a term in \lang\ into a pair of 
%syntactically % I don't like this word but temporarily here for continuity
separate, monostaged programs that correspond to the first and second stage computations in that term.

@@

More precisely, for any $\coltwo e A$ which reduces to a residual $q$ (via $\diatwo e q$),
$e$ splits into a syntactically separated program $\mathcal{P}$ 
(via the new relation $e \splittwosym \mathcal{P}$) which is
reduced to a residual $q'$ (via the new relation $\mathcal{P} \sepredtwosym q'$), which is the same as $q$ up to an equivalence.
By ``syntactically separated'' we mean that $\mathcal{P}$ 
has the form $\pipeS p l r$, where  
$p$ (called the {\em precomputation}) is a monostage term representing all the
first-stage computations in $e$, 
and $l.r$ (called the {\em resumer}) is a monostage term representing all the
second-stage computations in $e$.  
$\sepredtwosym$ is defined by a single rule, which evaluates the precomputation and substitutes the resulting value in for $l$ in the resumer:
\[
\infer{\sepredtwo {\pipeS p l r} {[b/l]r}} {\reduce p b}
\]
We call $b$ the {\em boundary value} resulting from the split since it represents the data communicated between the first and second stage computations.
A summary of these operations and the relationship between them is provided on the right side of \ref{fig:splittingSummary}.

Since terms at world \bbtwo\ can depend on terms at \bbonem\ (via
\texttt{prev}), splitting is also defined on \bbonem\ terms.  However,
defining correctness requirements for splitting \bbonem\ terms is
challenging because unlike \bbtwo\ terms (which reduce to monostage
residuals that can be directly compared to the monostage result of
$\sepredtwosym$), \bbonem\ terms reduce to partial values and residual
tables that represent computations at both stages.
%One complication arises from the fact that \lang\ terms at world \bbonem\ produce multistage output.
For example, the \bbonem\ term:
\begin{lstlisting}
(next {1+2}, grnd{3+4})
\end{lstlisting}
reduces (via $\redonesym$) to the residual table and partial value:
\begin{lstlisting}
[yhat |-> 1+2] (next {yhat}, grnd{7})
\end{lstlisting}

@@

Note that since $\coltwo e A$ reduces to an entirely second-stage
residual, its first-stage computations only exist to generate input
for second stage computations.  In contrast, since $\colmix e A$
reduces to a multi-stage residual table and partial value, its first
stage computations serve to both generate the value used as input for
the second stage {\em and also} produce a first-stage value.
Correspondingly, in the $\sepredtwosym$ rule, the precomputation $p$
generated via splitting reduces only to a boundary value which is
passed to the resumer; but in the $\sepredonesym$ rule, the combined
term $c$ reduces to a tuple containing both the first stage result $i$
and the boundary value $b$ provided as input to the resumer.  A
summary of the relationship between the results of directly evaluating
\lang\ and the results of splitting is provided at left in
\ref{fig:splittingSummary}.

In the following subsections we formally define the masking
transformation ($\vsplito$), then describe the algorithm for splitting
both \bbonem\ terms ($\splitonesym$) and \bbtwo\ terms
($\splittwosym$).

>>>>

In this section, we present a \emph{splitting algorithm} which
statically converts a term in \lang\ into a pair of separate,
monostaged terms that individually correspond to the first and second
stage computations of the \lang\ term. Due to important differences in
structure, we describe both the correctness condition and algorithm
for splitting of \bbonem\ terms separately from that of \bbtwo\ terms.
By construction, \bbonep\ terms are monostage and do not require
splitting.

@@

%More precisely
For any $\coltwo e A$ which reduces to a residual $q$ (via $\diatwo e q$),
$e$ splits into a syntactically separated program $\mathcal{P}$ 
(via the new relation $e \splittwosym \mathcal{P}$) which is
reduced to a residual $q'$ (via the new relation $\mathcal{P} \sepredtwosym q'$), which is the same as $q$ up to an equivalence.
By ``syntactically separated'' we mean that $\mathcal{P}$ 
has the form $\pipeS p l r$, where  
$p$ (called the {\em precomputation}) is a monostage term representing all the
first-stage computations in $e$, 
and $l.r$ (called the {\em resumer}) is a monostage term representing all the
second-stage computations in $e$.  
$\sepredtwosym$ is defined by a single rule, which evaluates the precomputation and substitutes the resulting value in for $l$ in the resumer:
\[
\infer{\sepredtwo {\pipeS p l r} {[b/l]r}} {\reduce p b}
\]
We call $b$ the {\em boundary value} resulting from the split since it represents the data communicated between the first and second stage computations.
A summary of these operations and the relationship between them is provided on the right side of \ref{fig:splittingSummary}.

%Since terms at world \bbtwo\ can depend on terms at \bbonem\ (via
%\texttt{prev}), splitting is also defined on \bbonem\ terms.  However,
Defining the correctness requirements for splitting \bbonem\ terms is
challenging because unlike \bbtwo\ terms (which reduce to monostage
residuals that can be directly compared to the monostage result of
$\sepredtwosym$), \bbonem\ terms reduce to partial values and residual
tables that represent computations at both stages.
%One complication arises from the fact that \lang\ terms at world \bbonem\ produce multistage output.
For example, the \bbonem\ term:
\begin{lstlisting}
(next {1+2}, grnd{3+4})
\end{lstlisting}
reduces (via $\redonesym$) to the residual table and partial value:
\begin{lstlisting}
[yhat |-> 1+2] (next {yhat}, grnd{7})
\end{lstlisting}

@@

Note that since $\coltwo e A$ reduces to an entirely second-stage
residual, its first-stage computations only exist to generate input
for second stage computations.  In contrast, since $\colmix e A$
reduces to a multi-stage residual table and partial value, its first
stage computations serve to both generate the value used as input for
the second stage {\em and also} produce a first-stage value.
Correspondingly, in the $\sepredtwosym$ rule, the precomputation $p$
generated via splitting reduces only to a boundary value $b$ which is
passed to the resumer; but in the $\sepredonesym$ rule, the combined
term $c$ reduces to a tuple containing both the first stage result $i$
and the boundary value $b$. A summary of the relationship between the
results of directly evaluating \lang\ and the results of splitting is
provided at left in \ref{fig:splittingSummary}.

The following subsections formally define the masking transformation
($\vsplito$), then describe the algorithm for splitting both
\bbonem\ terms ($\splitonesym$) and \bbtwo\ terms ($\splittwosym$).
